Simulating the Linux page cache in distributed applications with the WRENCH library

### 2020-12-09 paper_io_sim.tex
added comment

            Linux implements a disk cache called \textit{page cache}. 
            It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to avoid disk I/O by keeping data in memory 
            for later accesses, minimize I/O time since disk accesses are considerably 
            slower than memory accesses.
            When the kernel starts a read operation, it checks if required data is in page cache.
            If yes, data is read from memory. This is called a \textit{cache hit}. 
            Otherwise, data is read from disk and placed in page cache.
            For writing, page cache is implemented with three different strategies. 
            The first is \textit{direct~I/O} or \textit{no-write}, which means the page cache 
            is not involved in write operations. 
            The second is \textit{write-through}, in which data goes \textit{through} 
            page cache to disk. In this case, data is written to both page cache and disk. 
            In the third strategy, called \textit{writeback}, data are written directly 
            to disk as \textit{dirty} pages and these pages will be periodically flushed 
            to disk later by a \textit{writeback} process and then marked as not dirty. 
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes and perform a bulk write later.
            \cite{linuxdev3rd2010}.

            In writeback, one of the keys is the cache eviction mechanism, which is
            triggered when available memory is insufficient to remove pages 
            from page cache. 
            The term \textit{available memory} here includes both free memory and 
            the memory used but can be reclaimed by the kernel with flushing 
            and eviction. 
            Because dirty pages are data written to page cache but yet to be 
            written to disk, dirty pages must be kept in page cache and only clean pages 
            are available for eviction.
            If the clean pages are inadequate, dirty pages are flushed to disk, 
            marked as clean and available for eviction. 
            There is another type of flushing mechanism, which is 
            \textit{periodical flushing}, called periodically in predefined intervals. 
            Periodical flushing only flushes expired dirty pages, which remain dirty in 
            page cache longer than an expiration time configured in the kernel. 
            Different cache eviction algorithms have also been developed and proposed
            \cite{owda2014comparison}.
            In Linux kernel, a \textit{two lists strategy (LRU/2)} is implemented based on 
            the least-recently-used (LRU) scheme, with an \textit{active list} and 
            an \textit{inactive list}. 
            If accessed pages are not in the page cache, they are added to the inactive list. 
            If pages are accessed when residing in the inactive list, they will be moved from 
            inactive list to active list. 
            The lists are also kept balanced by moving pages from the active list
            to the inactive list when the active list grows too large.
            Thus, the active list only contains pages which are accessed more than once 
            and not evictable, while the inactive list includes pages accessed once only, 
            or pages accessed more than once but moved from the active list. 
            The lists work in LRU manner, pages are added to the tail and removed 
            from the top.

@@

            Many simulation frameworks and applications have been developed 
            in recent years to support studies in grids, cloud, HPC, peer-to-peer 
            networks, volunteer computers.                     
            These simulators are built with models of resources such as CPU, network, 
            storage with different approaches and levels of details in different simulators. 
            The detail level of the model causes a conflict between accuracy and scalability, 
            and these simulators usually trade of one for the other 
            \cite{casanova2014simgrid}. 
            For example, some simulators are built with packet-level network models 
            such as \textit{CloudSim} and \textit{iCanCloud}, or storage model that 
            simulates the operation of storage hardware in \textit{DiskSim} 
            \cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but at the same time lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, which allows the use of the framework 
            to simulate applications in multiple domains \cite{casanova2014simgrid}.

            Despite the fact that many sophisticated models for resources have been 
            integrated in simulators, and the page cache shows apparent impacts on I/O
            performance, the simulation model of page cache is still missing in most 
            of simulators.
            For example, SimGrid only simulates I/O operations with storage bandwidths 
            and capacity. 
            SIMCAN is a simulation framework that models page cache by storing data 
            accessed on disk in a block cache \cite{nunez2012simcan}. 
            Page cache is also modeled in iCanCloud simulator with a component in charge 
            of managing memory accesses and cached data \cite{nunez2012icancloud}, 
            but one of the issues is scalability due to its sophisticated model.
            Ultimately, non of these simulators provide any cache write strategy, 
            cache eviction mechanism with dirty/clean data and the page cache LRU lists. 
            Cache replacement policies is applied in \cite{xu2018saving} to simulate 
            in-memory caching, but for the study in energy consumption of multi-tier 
            heterogenous network.

            In this study, we chose SimGrid to implement our model of I/O with page cache 
            due to the balancing between accuracy and scalibility, the versatility allowing it 
            to be adopted in different domains. 
            In addition, it is an actively maintained open-scource framework with 
            modular design, which enables the framework to be extended with new models. 
            The main issue users have to face when using SimGrid is everything has to be 
            implemented from scratch due to its low-level API. 
            To ease this challenge, WRENCH, another open-source library, is built on top 
            of SimGrid to abstract the low-level logic and expose a higher abstraction 
            level API \cite{casanova2018wrench}. 
            We integrate our model in WRENCH and use the SimGrid API to interact with 
            simulated storage resource.
            This approach allows us to leverage existing features in WRENCH and SimGrid.
            Firstly, we can take advantage of the process control, synchronization of 
            resource consumption by concurrent processes provided by the simulation core 
            in SimGrid. 
            Secondly, we can reuse the existing SimGrid storage model with capacity 
            and bandwidth properties, which are similar to the characteristics of memory.
            And finally, WRENCH provides a high level abstraction, which help us drop 
            the burden of implementation.

>>>>

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, write-through enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends write-through
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes to perform a bulk write at a later time.
            \cite{linuxdev3rd2010}.

            Cache eviction and flushing strategies are integral to proper page cache functioning.
            Whenever space in memory becomes limited, either as a result of application memory
            or page cache use, page cache data may be evicted. Only data that 
            has been persisted to storage (clean pages) can be flagged for eviction and removed from
            memory. Written data that has not yet been persisted to disk (dirty data) must first
            be copied (flushed) to storage prior to eviction. When sufficient memory is
            being occupied, the flushing process is synchronous. However, even when 
            there is sufficient available memory, written data will be flushed to disk
            at a predefined interval through a process known as \textit{periodical flushing}.
            Periodical flushing only flushes expired dirty pages, which remain dirty in 
            page cache longer than an expiration time configured in the kernel. 
            Different cache eviction algorithms have also been developed and proposed
            \cite{owda2014comparison}.
            The Linux kernel uses a two-list strategy to flag pages for eviction.
            The two-list strategy is based on a least recently used (LRU) policy
            and uses an active and inactive list in its implementation.
            If accessed pages are not in the page cache, they are added to the inactive list. 
            Should pages located on the inactive list be accessed, they will be moved from 
            the inactive to the active list. 
            The lists are also kept balanced by moving pages from the active list
            to the inactive list when the active list grows too large.
            Thus, the active list only contains pages which are accessed more than once 
            and not evictable, while the inactive list includes pages accessed once only, 
            or pages that have be accessed more than once but moved from the active list. 
            Both lists operate using LRU eviction policies meaning the data that has
            not be accessed recently will be moved first.

@@

            Many simulation frameworks exist to describe the behaviours of applications
            running on the various infrastructures available to users, such as personal computers,
            clouds and clusters.
            These simulators are built with resource models such CPU, network and 
            storage, using various different approaches to describe the hardware.
            For example, there is \textit{CloudSim} and \textit{iCanCloud} for packet-level network models 
            and \textit{DiskSim} the simulates storage hardware functioning.

            Simulators must choose between scalability and accuracy when describing the
            hardware.\cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, allowing the use of the framework 
            to simulate applications in multiple domains. WRENCH, a workflow management
            system simulator based on the SimGrid framework, provides a high-level API
            for simulating workflows, thus facilitating the integration of new models.
            \todo{say more about wrench}

            Despite the fact that many sophisticated models for resources have been 
            integrated in simulators, and the page cache shows apparent impacts on I/O
            performance, the simulation model of page cache is still missing in most 
            of simulators.
            For example, SimGrid only simulates I/O operations with storage bandwidths 
            and capacity. 
            SIMCAN is a simulation framework that models page cache by storing data 
            accessed on disk in a block cache \cite{nunez2012simcan}. 
            Page cache is also modeled in the iCanCloud simulator through a component which
            manages memory accesses and cached data \cite{nunez2012icancloud}. However,
            the scalability of the iCanCloud simulator is limited by its sophisticated model.
            Ultimately, none of these simulators provide any writeback cache simulator nor
            cache eviction policys through LRU lists.
            Although cache replacement policies are applied in \cite{xu2018saving} to simulate 
            in-memory caching, their simulator is specific to energy consumption of multi-tier 
            heterogenous networks.

            In this study, we use WRENCH 
            to implement our page cache model. We chose WRENCH due to accurate yet scalable implementation, resulting
            from its SimGrid backend, as well as its high-level API.
            We have integrated our model in WRENCH and use the SimGrid API to interact with 
            simulated storage resources.
            This approach allows us to leverage existing features in WRENCH and SimGrid such as process control and synchronization of 
            resource consumption.
            We also leverage the existing SimGrid storage model with capacity 
            and bandwidth properties, which are similar to the characteristics of memory.
